MPA Gap Analysis: IUCN and AquaMaps

### 2016-02-22 mpa_analysis/wdpa_prep.Rmd
oh god did I finally get past this too-large-file bullshit?

# MPA gap analysis

Several steps to this:

* set up WDPA database in a raster format for easy use
* perform gap analysis for WDPA-MPA data base (which year?) vs AquaMaps 2014, with 50% threshold - replicate the original analysis
* reperform the analysis with latest/greatest WDPA-MPA database, and AquaMaps 2015, with 50% threshold and 0% threshold
* perform the analysis with latest/greatest WDPA-MPA database, and IUCN shapefiles

## set up WDPA raster

* direct extract method: extract WDPA polygons against half-degree cells, with proportion; value = year (or combo code)
    * all protected areas
    * 'marine' = 1
    * 'marine' = 1 and IUCN I - IV
    * clip WDPA to ocean regions, then IUCN I - IV.
    * NOTE: using extract() could result in double-counting - at least make sure total protected area per cell is not greater than 1.
        * maybe check any cells with protection > 50% and protected areas n > 1?
    * this method would be excruciatingly slow.
* to avoid double-counting for overlapping polygons - rasterize WDPA to fine scale first
    * less than 1km^2^ was mentioned in paper as resolution of MPA data.  Rasterize to 0.01 degree x 0.01 degree?  
        * this results in max cell size of 1.11 km x 1.11 km at the equator.
        * -180 to +180 longitude; min/max latitude use min/max from AquaMaps
    * WDPA_MPA layer: value = IUCN category (1-4, with NA = 7?) combined with marine flag and marine overlap flag?
        * first digit IUCN category; second digit marine flag; third digit marine overlap?
            * rasterize by IUCN category (unprotected, call NA?)
            * then rasterize again by marine field (0 or 1)
            * then rasterize land (or ocean) (0 or 1)
            * then add 'em up to construct full raster at 0.01 degree resolution
        * for year-to-year analysis, could recreate this layer for older years by filtering out newer protected areas
            * this allows for a cell's protection class to change over time?
            * could do this every, say, 5 years - to more quickly capture the trend across a span of, say, 25 years.
    * Cell area layer: use raster::area() to create a single layer of cell areas at this new resolution.
    * aggregate cells up to 0.5 degree cells, by protected area within each cell for each type of analysis
        * summarize to data.frame of loiczid, with area of protection (and total area?)?
        * create different protected-area layer for each protection scenario
* AquaMaps layer: use as is
* To analyze for each species:
    * Summarize total area and total protected area (for a given protection scenario) for the species
    * Include species extinction risk in there too?

``` {r set_up_base_raster}
### base raster will contain 0.01 degree cells from -180 to +180,
### and -90 to +90, with the associated area of each cell.
### Save it as a .tif in GitHub - about 50 
rast_base_file <- file.path(dir_git, 'mpa_analysis/rast_base.tif')

if(!file.exists(rast_base_file)) {
  ### import AquaMaps half-degree loiczid raster - for extents and such
  rast_loiczid_file   <- file.path(dir_git, 'data/rasters/loiczid_raster.grd')
  rast_loiczid        <- raster(rast_loiczid_file)
  names(rast_loiczid) <- 'loiczid'

  rast_empty <- raster(ext = extent(rast_loiczid), resolution = 0.01)
  rast_base  <- raster::area(rast_empty) ### takes a while...

  writeRaster(rast_base, rast_base_file, overwrite = TRUE)
} else {
  rast_base <- raster(rast_base_file)
}

@@

Dealing with rasters:

* how to ensure 1 km or better resolution?  WGS84 at 0.01 degree or Mollweide at 1 km
    * WGS84 --> 36000 * 18000 cells = 648,000,000 cells...
    * Mollweide --> 
* options for creating rasters
    * use arcpy as in LSP... global Mollweide raster of WDPA is about 50 MB; includes .tif and supporting files.  This is .5 km raster.  How is it so small?
    * use gdal_rasterize seems to create multi-GB rasters; .tif only
* why rasters?  avoid double-counting for overlapping parks.  WDPA is most important, but since that is in raster and AM is raster, easiest to keep it all in raster
    * WDPA - capture park classification and year and marine flag
    * spatial marine - capture whether cell includes ocean or not - clip to max N and S of WDPA?
        * xMin,yMin -180.007,-72.95 : xMax,yMax 180,83.73
* Use this R script to source arcPy script modified from LSP - need to run it on a PC with appropriate GDAL installation, most likely

``` {r set_up_wdpa_raster}

### import WDPA database polygons from saved .shp
dir_wdpa   <- file.path(dir_N, 'git-annex/globalprep/_raw_data/wdpa_mpa/WDPA_Jan2015_Public')
lyr_wdpa <- 'WDPA_poly_Jan2015'
poly_wdpa <- readShapePoly(file.path(dir_wdpa, lyr_wdpa), 
                           proj4string = crs_wgs84)

prot_code_df <- data.frame('IUCN_CLASS' = c('I', 'IIa', 'IIb', 'III', 'IV', 'V', 'V', NA),
                           'prot_code'  = c( 1,    2,     2,     3,     4,   5,   6,   7))
poly_wdpa@data <- poly_wdpa@data %>%
  left_join(prot_code_df, by = 'IUCN_CLASS')

poly_mpa <- poly_wdpa[poly_wdpa@data$MARINE == 1, ]
poly_mpa <- poly_wdpa[poly_wdpa@data$prot_code <= 4, ]

>>>>

``` {r set_up_wdpa_mpa_polys}

### import WDPA database polygons from saved .shp
dir_wdpa   <- file.path(dir_N, 'git-annex/globalprep/_raw_data/wdpa_mpa/WDPA_Jan2015_Public')
lyr_wdpa    <- 'WDPA_poly_Jan2015'
lyr_mpa     <- 'WDPA_poly_Jan2015_mpa'
lyr_mpa_mar <- 'WDPA_poly_Jan2015_mpa_mar'

if(!file.exists(file.path(dir_wdpa, paste(lyr_mpa, '.shp', sep = '')))) {
  poly_wdpa <- readOGR(dsn = dir_wdpa, layer = lyr_wdpa, stringsAsFactors = FALSE)

#   nrow(poly_wdpa@data)
#   poly_wdpa@data$IUCN_CAT %>% unique()

  prot_code_df <- data.frame('IUCN_CAT'  = c('I', 'Ia', 'Ib', 'II', 'III', 'IV', 'V', 'VI'),
                             'prot_code' = c(  1,   1,    1,    2,     3,    4,   5,    6))
# poly_wdpa_save <- poly_wdpa
  poly_wdpa@data <- poly_wdpa@data %>%
    left_join(prot_code_df, by = 'IUCN_CAT') %>%
    mutate(prot_code = ifelse(is.na(prot_code), 7, prot_code))

  poly_mpa <- poly_wdpa[poly_wdpa@data$STATUS == 'Designated', ]
  poly_mpa <- poly_mpa[poly_mpa@data$prot_code <= 4, ]

  poly_mpa_marine <- poly_mpa[poly_mpa@data$MARINE == 1, ]

  writeOGR(poly_mpa, dsn = dir_wdpa, layer = lyr_mpa,
           driver = 'ESRI Shapefile')
  writeOGR(poly_mpa_marine, dsn = dir_wdpa, layer = lyr_mpa_mar,
           driver = 'ESRI Shapefile')
}

poly_mpa <- readOGR(dir_wdpa, lyr_mpa, stringsAsFactors = FALSE)
### create simplified polygon
polys_mpa_area <- lapply(poly_mpa@polygons, function(x) x@area)
poly_mpa1 <- poly_mpa[polys_mpa_area > 1e-5, ]
poly_mpa1 <- poly_simplify(poly_mpa1, thresh = 1e-5, verbose = TRUE)

### sort by STATUS_YR, descending, so gdal_rasterize will assign newer
### values then overwrite (where applicable) with older values
poly_mpa2 <- poly_mpa1[order(z <- poly_mpa1@data$STATUS_YR, decreasing = TRUE), ]

writeOGR(poly_mpa1, dsn = dir_wdpa,
         layer = paste(lyr_mpa, '1', sep = ''),
         driver = 'ESRI Shapefile')

}
```

unique(poly_mpa@data$STATUS_YR)
table(poly_mpa@data$STATUS_YR)
hist(poly_mpa@data$STATUS_YR[poly_mpa@data$STATUS_YR > 0])
summary(poly_mpa@data$STATUS_YR[poly_mpa@data$STATUS_YR > 0])

``` {r rasterize_that_shit}
dir_rast <- file.path(dir_anx, 'explore_am_v_iucn/mpa_rasters')

### Get source from WDPA_MPA data location
src <- path.expand(file.path(dir_wdpa, paste(lyr_mpa, '1.shp', sep = '')))
### gdal_rasterize to temp directory
tmp_dst <- file.path(dir_rast, 'tmp', paste(lyr_mpa, '1.tif', sep = '')) %>%
  path.expand
### then writeRaster to final directory, to save smaller file
fnl_dst <- file.path(dir_rast, paste(lyr_mpa, '1.tif', sep = '')) %>%
  path.expand

### Set extents and resolution
ext  <- c(-180, -90, 180, 90)
reso <- c(.01, .01)

ptm <- proc.time()
tmp_rast <- gdalUtils::gdal_rasterize(
  src_datasource = src,
  dst_filename   = tmp_dst, 
    # destination for output
  a = 'STATUS_YR', 
    # the attribute in the shapefile to be assigned to the cell values
  te = ext, 
    # extents for output raster
  tr = reso,  
    # resolution for output raster
  a_nodata = NA, 
    # nodata value for raster; otherwise they will be filled in with zeroes
  output_Raster = TRUE,
    # return output as a RasterBrick? 
  verbose = TRUE)
message('Rasterized ', src, '; \n  ', round((proc.time() - ptm)[3], 2), ' seconds')

ptm <- proc.time()
raster::writeRaster(tmp_rast, fnl_dst, overwrite = TRUE)
message('Write ', src, '\n  to ', fnl_dst, '; \n  ', round((proc.time() - ptm)[3], 2), ' seconds')
remove(tmp_rast)
unlink(tmp_dst)

@@

``` {r simplify_polys_function}
poly_simplify <- function(poly_set, thresh, verbose = FALSE) {
  if(class(poly_set) == 'list') poly_set <- poly_set[[1]]
#   if(verbose) {
#     message('Counting subpolys in poly set:')
#     poly_counts <- sapply(poly_set@polygons, function(x) length(x@Polygons))
#     print(poly_counts)
#   }

  ### filter out tiny sub-polygons -----
  ### create area list for filtering.  Area will be in square degrees... 
  polys_area <- lapply(poly_set@polygons, function(x) sapply(x@Polygons, function(y) y@area))

  if(verbose) {
    message('Polygon area quantiles:')
    print(quantile(unlist(polys_area)))
  }

  ### check to make sure threshold 
#   if(verbose) {
#     message('Smallest "main" subpolygon, i.e. smallest EEZ, in this set:')
#     poly_small <- min(unlist(lapply(polys_area, function(x) (max(x)))))
#     if(poly_small < thresh) message('Smallest poly smaller than threshold value of ', thresh)
#     for (j in 1:length(polys_area)) {
#       message(sprintf('rgn %s: largest polygon %.5f (%s)', poly_set@data$rgn_id[j], 
#                       max(polys_area[[j]]), poly_set@data$rgn_nam[j]))
#     }
#   }

  mainPolys <- lapply(polys_area, function(x) which(x > thresh))

  for(i in 1:length(mainPolys)) {
    if(length(mainPolys[[i]]) >= 1 && mainPolys[[i]][1] >= 1) {
      poly_set@polygons[[i]]@Polygons  <- poly_set@polygons[[i]]@Polygons[mainPolys[[i]]]
      poly_set@polygons[[i]]@plotOrder <- 1:length(poly_set@polygons[[i]]@Polygons)
    }
  }

  plot_order_max <- max(poly_set@plotOrder)
  poly_list_length <- length(poly_set@polygons)
  message('Plot order max: ', plot_order_max, '; length poly list: ', poly_list_length)

  if(plot_order_max != poly_list_length) {
    message('NOT SAME LENGTH?!')
    stop()
    # poly_set@plotOrder <- 1:length(poly_set@polygons)
    # is plotOrder necessary to be continuous?  that would explain this line.
  }

  if(verbose) {
#     message('Counting subpolys in modified poly set:')
#     poly_counts <- sapply(poly_set@polygons, function(x) length(x@Polygons))
#     print(poly_counts)

    polys_area_new <- lapply(poly_set@polygons, function(x) sapply(x@Polygons, function(y) y@area))

    message('Polygon area quantiles after adjustment:')
    print(quantile(unlist(polys_area_new)))
  }

  return(poly_set)
}
